\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-24 paper/experiments.tex
Update perf discussion
paper: intro feedback from martin

Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for
users based on their income data. In Query~\ref{q2}, we compare cholesterol levels for
individuals with low, medium, and high incomes spectrum and above a certain weight. In
Query~\ref{q3}, we extract the maximum blood lead levels for children under the age of 6
years. In Query~\ref{q4}, we calculate the average systolic blood pressure, by gender, for
subjects with a body mass index indicating obesity. In Query~\ref{q5}, we extract the age,
gender, triglyceride levels, and waist circumference from individuals with high triglyceride
levels.

Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for
survey participants, based on their bootcamp attendance. In Query~\ref{q7} estimates the
average age of women from the United States who participated. In Query~\ref{q8}, we
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a
reference table provided by the World Bank which summarizes GDP per-capita across various
countries\cite{worldbank-data}. The query calculates the average GDP per-capita for
participants based on whether or not they attended a bootcamp. Finally, in Query~\ref{q10},
we identify individuals from countries with GDP per-capita of \$5000 or less who earn
\$100,000 or more post-bootcamp.

@@

%\todobox{actually run experiments and rewrite this section}{
All experiments were run on a single Amazon Web Services EC2 {\tt c4.xlarge} instance, with
four 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory, on Debian Linux.

@@

In many real-world cases, applying the imputation step at the base table is prohibitive.
To illustrate the increasing difficulty of such an approach as datasets scale, we used a simple
aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}).
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 75 minutes. In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.

\begin{figure}
\begin{lstlisting}
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
\caption{A simple aggregate query on ACS data. Imputing the base table is prohibitive.
    \ProjectName's quality-optimized and performance-optimized queries execute in 7 seconds
    and 1 second, respectively.
}
\label{query-acs}
\end{figure}

>>>>

Queries 1-5 are on the CDC data. In Query~\ref{q1}, we calculate the average height for users based on their income data. In Query~\ref{q2},
we compare cholesterol levels for individuals at either tail of the income
spectrum and above a certain weight. In Query~\ref{q3},
we extract the maximum blood lead levels for children under the age of 6 years. In Query ~\ref{q4}, we calculate the average systolic blood pressure, by gender, for subjects with a body mass index indicating obesity. In Query ~\ref{q5}, we extract the age (in months), creatine level, and head circumference, for subjects with high creatine levels.

% Perhaps more school makes your head bigger :)

Queries 6-10 are on the FCC data. In Query~\ref{q6}, we calculate the average income for survey participants,
based on their bootcamp attendance. In Query~\ref{q7} estimates the average
age of women from the United States who participated. In Query~\ref{q8}, we 
calculate the average amount of money survey participants with student debt destined to
learning based on their school degree. In Query~\ref{q9}, we join the FCC data with a reference table provided
by the World Bank which summarizes GDP per Capita across various
countries\cite{worldbank-data}. The query calculates the average GDP-per-capita for participants based on whether or not they
attended a bootcamp. Finally, in Query ~\ref{q10}, we identify individuals from countries with GDP per-capita of \$5000 or
less who earn \$100,000 or more post-bootcamp.

@@

%\todobox{actually run experiments and rewrite this section}{
All experiments were run on a single AWS EC2 c4.xlarge instance, with 4 2.9 GHz Intel Xeon E5-2666 v3 virtual CPUs and 7.5 GiB of main memory on Debian Linux.

@@

% TODO MJS
In many cases, applying the imputation step at the base table simply is not an option. To illustrate the increasing difficulty of such an approach
as datasets scale, we used a simple aggregate query over the ACS dataset in our final experiment(Figure~\ref{query-acs}). Applying the imputation operation to the base table
is extremely expensive, in our case the process did not terminate within 20 minutes. In contrast, \ProjectName{} executes a quality-optimized version
in 9 seconds and a runtime-optimized version in 1 second. This highlights the potential benefit of using our system for early data exploration.

\begin{figure}
\begin{lstlisting}
SELECT AVG(c0) FROM acs_dirty;
\end{lstlisting}
\caption{A simple aggregate query on ACS data. Imputing the base table is undesirable (did not terminate within 20min). \ProjectName's
quality-optimized and performance-optimized query execute in 9 seconds and 1 second, respectively.
}%TODO MJS
\label{query-acs}
\end{figure}

### 2017-02-24 paper/introduction.tex
paper: working on optimality proof
paper: intro feedback from martin

% Solution
To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on-the-fly, during query execution.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.

% Prior work
There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
Prior work in statistics has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
In contrast, \ProjectName{} does not focus on the imputation algorithm itself but rather on the planning involved in its use in non-trivial query plans.

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set and get good results, even if that data has missing values.

% Background
Traditional imputation methods work over the entire data set and replace all missing values.
Although this solves the problem of missing data, running a sophisticated imputation algorithm over a large data set can be very expensive.
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.

% Scope
\ProjectName{} is designed to enable data exploration, by allowing analysts to run their queries and get good results without dealing with imputation.
However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
This paper makes the following contributions:
\begin{itemize}
\item \textbf{Relational algebra with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present a novel query planning algorithm which jointly optimizes for running time and the quality of the results.
  It does so by maintaining the set of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
\end{itemize}

>>>>

% Solution
To handle this problem, users may manually clean their dataset by performing some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
We have developed a system called \ProjectName{} with a novel query optimization algorithm which allows imputation to occur on the fly, during query execution.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.

% Prior work
There is relevant prior work in three broad areas: statistics, databases, and time-series forecasting.
Prior work in statistics has focused on imputation quality~\cite{burgette2010multiple} and runtime cost~\cite{akande2015empirical}.
Prior work in databases has a established simple semantics for handling tuples with missing values~\cite{codd1973understanding,grant1977null}.
Prior work in time-series forecasting has incorporated forecasting extensions into domain-specific databases and characterized their behavior~\cite{parisi2011embedding,parisi2013temporal,duan2007processing}.
\ProjectName{} contributes a novel approach to planning data imputation on a per query basis, allowing users to use standard SQL on real-world data sets with missing values
and express preferences in quality of data imputation versus running time.

% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive.
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
Finally, for many queries, it is not necessary to clean the entire data set, and a significant speed up can be obtained by cleaning only the necessary parts.
In contrast to existing systems, \ProjectName{} exploits this observation and avoids imputing over the entire data set, as done with traditional methods. \ProjectName{}
carefully plans the placement of different imputation operations throughout the query plan, resulting in a 
significant speedup in query execution. The focus on of our work is not on the imputation algorithm but rather on the planning involved in its use in non-trivial query plans.

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
queries that execute in a fraction of the time it takes to impute base tables and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the base-table imputation approach, as indicated by low symmetric-Mean-Absolute-Percentage-Errors in our empirical results
over real-world data sets (see Sec.~\ref{sec:experiments} for details).

% Scope
\ProjectName{} is designed to enable early data exploration, by allowing analysts to run their queries without explicitly dealing with imputation.
However, we optimize query plans with respect to a heuristic measure of imputation quality, and provide no hard guarantees about the accuracy of the results.

\subsection{Contributions}
This paper makes the following contributions:
\begin{itemize}
\item \textbf{Relational algebra with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  Impute fills in missing data values using a regression model (CE-CART)~\cite{burgette2010multiple}.
  Drop simply drops tuples which contain null values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present a novel query planning algorithm which jointly optimizes for running time and the quality of the results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
\item \textbf{Proof of optimality:}
 We prove that our cost-model and planning algorithm produces a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

\ProjectName{} leverages the combined contributions to plan fast queries on data with missing values, while producing comparable results to traditional approaches.
This provides the opportunity to increase productivity for users who frequently encounter missing values in their data.
